Eleven years of student replication projects provide evidence on the correlates of replicability in psychology

### 2023-04-27 paper/manuscript.tex
work on abstract and intro

\begin{abstract}
A cumulative science relies on knowing how much to trust published findings. Large-scale replication studies provide a measure of how reliable the published literature in a field is by determining what fraction of studies replicate. Because of the effort involved in conducting a large number of replications, few large-scale replications are available. We add 176 replication projects conducted by students of experimental studies primarily in psychology. Roughly half the studies were replicated successfully, and we report how features of the studies and the replications influenced the replication outcome.

@@

A cumulative science requires foundational empirical results that can be build upon in theories and future experiments. These load-bearing results need to be reliable; when they are not robust, attempts to build on them waste resources, trying to find explanations and effects where there may be none.

Scientists, policymakers, and the public often treat publication as a signal that a result is (likely to be) true. This trust in published findings leads scientists to attempt to build on them, but failures to build on them are explained away and rarely published, leaving the original results uncontested in the literature, while failures accumulate in private.

Publicized failures to replicate called into question the tacit assumption that most of the literature was robust and would replicate. Certain prominent findings turned out to not replicate at all {[}ex. terror management theory; Klein et al. (\protect\hyperlink{ref-klein2022}{2022}){]}, and a large scale replication project pegged the replication rate for findings in top-tier psychology journals at around 40\% (\protect\hyperlink{ref-openscienceconsortium2015}{Consortium 2015}).

Few large scale replications have been conducted in psychology, due to the resources required to run large numbers of replication studies. We are aware of three. RP:P sampled roughly 100 studies from articles published in three top psychology journals in 2008 and found a overall replication rate of around 40\% (\protect\hyperlink{ref-openscienceconsortium2015}{Consortium 2015}). ManyLabs investigated heterogeneity using short target studies that compared between two conditions each, but these studies are not representative of the psychology literature as a whole. Across Many Labs 1-3, 29 of 51 target effects (57\%) replicated (\protect\hyperlink{ref-klein2014}{Klein et al. 2014}, \protect\hyperlink{ref-ebersole2016}{Ebersole et al. 2016}, \protect\hyperlink{ref-klein2018}{Klein et al. 2018}). Camerer et al. (\protect\hyperlink{ref-camerer2018}{2018}) replicated all 21 behavioral social science studies from Nature and Science from 2010-2015 that were feasible; the replication rate was around 60\%. These roughly 170 replications of psychology studies are the primary empirical results on which discussions of replication rates in psychology are based.

A replication rate around 50\% runs counter to the prior assumption that most findings in the literature was robust enough to support cumulative science, and raises the question: What makes the replication rates in psychology so low?

@@

Experimental approaches attempt to intervene on the low replication rate, by seeing if changing certain factors affects whether studies replicate. Protzko et al. (\protect\hyperlink{ref-protzko2020}{2020}) showed, across 16 studies, that better methodological practices, such as transparency, large sample sizes, and confirmatory tests, led to replication rates that matched theoretical expectations based on the effect sizes and sample sizes, with replication effect sizes comparable with the original. Many Labs 5 examined how adding expert advice to a replication process might intervene to improve the replicate rate, and found that it mostly didn't, at least on the 10 studies they looked at (\protect\hyperlink{ref-ebersole2020}{Ebersole et al. 2020}). These types of experiments are valuable for testing potential causes of non-replication, but they don't scale well do to expense. Additionally, not all potential influences on non-replication are experimentally manipulable.

Correlational approaches looking for predictors of replicability are the most popular. Prediction markets and elicitations have established that people can predict what studies will replicate above chance (\protect\hyperlink{ref-dreber2015}{Dreber et al. 2015}, \protect\hyperlink{ref-camerer2018}{Camerer et al. 2018}, \protect\hyperlink{ref-forsell2019}{Forsell et al. 2019}, \protect\hyperlink{ref-hoogeveen2019}{Hoogeveen et al. 2019}), but have not identified many concrete predictors that differentiate replications from non-replications. Machine learning approaches CITE similarly show predictive signal, but do not provide explainable predictor variables. (\protect\hyperlink{ref-opensicenceconsortium2015}{\textbf{opensicenceconsortium2015?}}) looked for simple correlates of replicability in the RP:P sample and found that studies in cognitive psychology (as opposed to social psychology) and studies with larger effect sizes and smaller p-values were more likely to replicate. Correlations approaches depend on data from replications, generally drawing heavily from the same small set of data points. In particular, the RP:P dataset itself is much discussed and reanalyzed (\protect\hyperlink{ref-anderson2016}{Anderson et al. 2016}, \protect\hyperlink{ref-etz2016}{Etz \& Vandekerckhove 2016}, \protect\hyperlink{ref-gilbert2016}{Gilbert et al. 2016}, \protect\hyperlink{ref-patil2016}{Patil et al. 2016}) to the point that much of what we think we know about replicability may be overfit to the 100 studies included in in RP:P.

@@

Further, the methods for estimating this quantity have been theoretically problematic. Sampling schemes for prior work typically do not reflect an entirely random sample from the literature; instead they sample from specific journals where results may be of more interest and adjust the sample for feasibility concerns. These are reasonable sampling choices, but they undermine the claim that the estimand is the level of ``truth'' in the literature as a whole. Sampling truely at random from the literature may not even be desirable, as arguably a literature will succeed if useful discoveries come out of it, not if random findings are true (\protect\hyperlink{ref-wilson2020}{Wilson et al. 2020}). How much we care about whether a study is ``trustworthy'' is not uniformly distributed across the literature.

In contrast, we have explicitly pursued a different estimand: the probability that a researcher, on selecting a finding of interest from the literature, can successfully achieve a result satisfactorily close enough to the original that they can build on it in their own work, with all the necessary compromises to the methods and sample of the original that may be required by the constraints of the situation.

This framing of replicability matches our methodology.

Rather than sampling at random from some parts of the literature; our sample of studies is selected based on what studies students were interested in and wanting to replicate, with some filtering for feasibility; this sampling reflects how scientists choose what studies to build on: those that are interesting and relatively doable given methodological and budgetary constraints.

We use a subjective replication score as our primary metric. Whether one feels confident in the results of a study given a replication is not always dependent on only one outcome measure (ex. interaction term) and particularly not dependent on only one statistical comparison between the two studies (ex. replication is p\textless.05 same direction as original). This avoids bright line distinctions and accomodates the range of outcome measures in our diverse set of studies.

All our replications were conducted under short time scales and relatively small budgets, which mimics the constraints many scientists are under when starting new projects. From the point of view of cumulative science, researchers want to know if they can get paradigms to work under their real-world constraints. The same issue that may cause a study not to replicate under constrained circumstances (despite hypothetically replicating under more favorable circumstances such as expert administration or larger budgets) will also plague attempts to build off those studies in constrained circumstances. Thus, we believe it is relevant to estimate replicability as done by a graduate student with limited resources.

@@

Nearly all replications studies were conducted online, but original studies were split between using in-person and online recruitment. Replications that switched to online were less likely to replicate than those that had the same modality as the original (generally both online, in a few cases both in-person). While online studies in general show comparable results to studies conducted in person (\protect\hyperlink{ref-crump2013}{\textbf{crump2013?}})TODO MORE CITATIONS, switching the modality does decrease the closeness of the replication, and some studies done in person may not have been well adapted (ex. inductions may be weaker or attention checks inadequate to the new sample).

@@

\begin{figure}[ht]
\includegraphics[width=0.9\linewidth]{manuscript_files/figure-latex/mod-results-1} \caption{Coefficient estimates and uncertainty from a model predicting subjective replication scores from the full dataset.}\label{fig:mod-results}
\end{figure}

@@

The replication rate in psychology is to the best estimates, somewhere around 50\%, which is quite low. Many reasons for this low replication rate have been hypothesized, but studying it either experimentally or correlationally requires doing a number of replication experiments, which can be a lot of work. Here we took advantage of 11 years of graduate student replication projects to look at correlational predictors of replication in a previously-unused dataset.

@@

Our results are limited by the number and quality of the studies we included. These studies are not necessarily representative of the studies of interest to psychologists as a whole. The replication studies were not designed from the start with this analysis in mind, so analysis is limited by the choices and reporting used at the time of replications. TODO what are the limitations we want?

TODO quick wrap up

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-anderson2016}{}}%
Anderson CJ, Bahník Š, Barnett-Cowan M, Bosco FA, Chandler J, Chartier CR, Cheung F, Christopherson CD, Cordes A, Cremata EJ, Della Penna N, Estel V, Fedor A, Fitneva SA, Frank MC, Grange JA, Hartshorne JK, Hasselman F, Henninger F, Hulst M van der, Jonas KJ, Lai CK, Levitan CA, Miller JK, Moore KS, Meixner JM, Munafò MR, Neijenhuijs KI, Nilsonne G, Nosek BA, Plessow F, Prenoveau JM, Ricker AA, Schmidt K, Spies JR, Stieger S, Strohminger N, Sullivan GB, Aert RCM van, Assen MALM van, Vanpaemel W, Vianello M, Voracek M, Zuni K (2016) Response to {Comment} on {``{Estimating} the reproducibility of psychological science.''} \emph{Science} \textbf{351}:1037--1037. doi:\href{https://doi.org/10.1126/science.aad9163}{10.1126/science.aad9163}

@@

\leavevmode\vadjust pre{\hypertarget{ref-openscienceconsortium2015}{}}%
Consortium OS (2015) \href{https://www.science.org/doi/full/10.1126/science.aac4716?casa_token=IJ35TwwlcjsAAAAA\%3AqiP68QbVAHleIg9zD3WugKWuV6Oa5rswS0VQnDsCq5I14ME4WIQabNGVD_T6SBSuAt6voVHNnWc0sw}{Estimating the reproducibility of psychological science}. \emph{Science}

@@

@@

>>>>

\begin{abstract}
A cumulative science requires foundational empirical results that are robust enough to support theories and future experiments. Prominent failed replications of well-known published findings cast doubt on the psychological literature, and large scale replication projects suggested that the overall replication rate in psychology might be between 40-60\%. Large-scale replications are arduous, and only around 170 replications have been conducted as part of a large-scale replication project. This dearth of data points limits estimations of replicability and analyses of potential correlates of replicability. Using a new sample of 176 replications conducted as part of a course, we estimate replicability of these studies that were of interest to graduate students. We look at statistical and experimental-design features of studies as potential predictors of replication. We find that within-study designs and large original effect sizes positively correlate with replicability.

@@

A cumulative science requires foundational empirical results that are robust enough to support theories and future experiments. If scientists attempt to build cumulative science on top of results that are unreliable and cannot be achieved consistently, they waste resources looking for results where there may be none. This is frustrating for individual scientists, but also damaging to the enterprise of science overall.

Scientists, policymakers, and the public often treat publication in a scientific journal as a signal that a result is (likely to be) true. This trust in published findings has led scientists to build on findings without verifying them first, but failures were often explained away and rarely published, leaving the original results uncontested in the literature, while failures to achieve the same outcomes accumulated in private.

During the replication crisis, publicized failures to replicate published findings called into question the tacit assumption that most of the literature was robust and would replicate. Certain prominent findings failed to replicate in multi-site replication attempts (ex. terror management theory, \protect\hyperlink{ref-klein2022}{Klein et al. 2022}, ego-depletion, \protect\hyperlink{ref-hagger2016}{Hagger et al. 2016}), and a large-scale replication project pegged the replication rate for findings in top-tier psychology journals at around 40\% (\protect\hyperlink{ref-openscienceconsortium2015}{Open Science Consortium 2015}).

How replicable is psychology as a whole? Few large scale replications have been conducted in psychology, due to the resources required to run large numbers of replication studies. We are aware of three. RP:P sampled roughly 100 studies from articles published in three top psychology journals in 2008 and found a overall replication rate of around 40\% (\protect\hyperlink{ref-openscienceconsortium2015}{Open Science Consortium 2015}). ManyLabs investigated heterogeneity using short target studies that compared between two conditions each, but these studies are not representative of the psychology literature as a whole. Across Many Labs 1-3, 29 of 51 target effects (57\%) replicated (\protect\hyperlink{ref-klein2014}{Klein et al. 2014}, \protect\hyperlink{ref-ebersole2016}{Ebersole et al. 2016}, \protect\hyperlink{ref-klein2018}{Klein et al. 2018}). Camerer et al. (\protect\hyperlink{ref-camerer2018}{2018}) replicated all 21 behavioral social science studies from Nature and Science from 2010-2015 that were feasible; the replication rate was around 60\%. These roughly 170 replications of psychology studies are the primary empirical results on which discussions of replication rates in psychology are based.

A replication rate around 50\% runs counter to the prior assumption that most findings in the literature were robust enough to support cumulative science, and it raises the question: What makes the replication rates in psychology so low?

@@

Experimental approaches attempt to intervene on the low replication rate, by changing certain factors that may affect whether studies replicate. Protzko et al. (\protect\hyperlink{ref-protzko2020}{2020}) showed, across 16 studies, that better methodological practices, such as transparency, large sample sizes, and confirmatory tests, led to replication rates that matched theoretical expectations based on the effect sizes and sample sizes, with replication effect sizes comparable with the original. Many Labs 5 examined how adding expert advice to a replication process might increase the replication rate, and found that it mainly did not for the 10 studies they looked at (\protect\hyperlink{ref-ebersole2020}{Ebersole et al. 2020}). These types of experiments are valuable for testing potential causes of non-replication, but they don't scale well due to expense. Additionally, not all potential influences on non-replication are experimentally manipulable.

Correlational approaches looking for predictors of replicability are the most popular. Prediction markets and elicitations have established that people can predict what studies will replicate above chance (\protect\hyperlink{ref-dreber2015}{Dreber et al. 2015}, \protect\hyperlink{ref-camerer2018}{Camerer et al. 2018}, \protect\hyperlink{ref-forsell2019}{Forsell et al. 2019}, \protect\hyperlink{ref-hoogeveen2019}{Hoogeveen et al. 2019}), but have not identified concrete predictors that differentiate replications from non-replications. Using machine learning approaches, Altmejd et al. (\protect\hyperlink{ref-altmejd2019}{2019}) found that sample size, effect size, and simple effects were predictive of replication. Open Science Consortium (\protect\hyperlink{ref-openscienceconsortium2015}{2015}) looked at simple correlates of replicability in the RP:P sample and found that studies in cognitive psychology (as opposed to social psychology) and studies with larger effect sizes and smaller p-values were more likely to replicate. Correlational approaches depend on data from replications, generally drawing heavily from the same small set of data points. In particular, the RP:P dataset itself is much discussed and reanalyzed (\protect\hyperlink{ref-anderson2016}{Anderson et al. 2016}, \protect\hyperlink{ref-etz2016}{Etz \& Vandekerckhove 2016}, \protect\hyperlink{ref-gilbert2016}{Gilbert et al. 2016}, \protect\hyperlink{ref-patil2016}{Patil et al. 2016}) to the point that much of what we think we know about replicability may be overfit to the 100 studies included in in RP:P.

@@

Further, the methods for estimating this quantity have been theoretically problematic. Sampling schemes for prior work typically do not reflect an entirely random sample from the literature; instead they sample from specific journals where results may be of more interest and adjust the sample for feasibility concerns. These are reasonable sampling choices, but they undermine the claim that the estimand is the level of ``truth'' in the literature as a whole. Sampling truly at random from the literature may not even be desirable, as arguably a literature will succeed if useful discoveries come out of it, not if random findings are true (\protect\hyperlink{ref-wilson2020}{Wilson et al. 2020}). The importance of a study being replicable is not uniformly distributed across the literature.

In contrast to prior approaches, we have explicitly pursued a different estimand: the probability that a researcher, on selecting a finding of interest from the literature, can successfully achieve a result satisfactorily close enough to the original that they can build on it in their own work, with all the necessary compromises to the methods and sample of the original that may be required by the constraints of the situation.

Our methodology is reflective of this framing of replicability as being the probability of a studying being able to support cumulative science. Rather than sampling at random from some parts of the literature; our sample of studies is selected based on what studies students were interested in and wanting to replicate, with some filtering for feasibility; this sampling reflects how scientists choose what studies to build on: those that are interesting and relatively doable given methodological and budgetary constraints.

We use a subjective replication score as our primary metric. Whether one feels confident in the results of a study given a replication is not always dependent on only one outcome measure (ex. interaction term) and particularly not dependent on only one statistical comparison between the two studies (ex. replication is p\textless.05 same direction as original). This metric avoids bright line distinctions and accommodates the range of outcome measures in our diverse set of studies.

All our replications were conducted under short time scales and relatively small budgets, which mimics the constraints many scientists are under when starting new projects. From the point of view of cumulative science, researchers want to know if they can get paradigms to work under their real-world constraints. The same issue that may cause a study not to replicate under constrained circumstances (despite hypothetically replicating under more favorable circumstances such as expert administration or larger budgets) will also plague attempts to build off those studies in constrained circumstances. Thus, we believe it is relevant to estimate replicability under the limitations of real world resource and expertise constraints.

@@

Nearly all replications studies were conducted online, but original studies were split between using in-person and online recruitment. Replications that switched to online were less likely to replicate than those that had the same modality as the original (generally both online, in a few cases both in-person). While online studies in general show comparable results to studies conducted in person (\protect\hyperlink{ref-crump2013}{Crump et al. 2013}), switching the modality does decrease the closeness of the replication, and some studies done in person may not have been well adapted (ex. inductions may be weaker or attention checks inadequate to the new sample).

@@

\begin{figure}[ht]
\includegraphics[width=0.9\linewidth]{manuscript_files/figure-latex/mod-results-1} \caption{Coefficients and uncertaintly estimates from a model of all predictors (N=112, shown in blue) and non-statistical predictors (N=176, shown in red) predicting subjective replication scores as the dependent variable. }\label{fig:mod-results}
\end{figure}

@@

The replication rate in psychology is, to the best estimates, somewhere around 50\%, which is quite low. Many reasons for this low replication rate have been hypothesized, but studying it either experimentally or correlationally requires doing a number of replication experiments, which can be a lot of work. Here we took advantage of 11 years of graduate student replication projects to look at correlational predictors of replication in a previously-unused dataset.

@@

Our results are limited by the number and quality of the studies we included. These studies are not necessarily representative of the studies of interest to psychologists as a whole. The replication studies were not designed from the start with this analysis in mind, so analysis is limited by the choices and reporting used at the time of replications. We reiterate that given our sample, we are able to estimate the correlates of studies replicating when the replications are conducted online, by graduate students with limited time and budget. This estimand is not the same as whether they would replicate when done by large labs with access to more money and experts. This estimand does not represent whether or not the original study was ``true''.

TODO conclusion

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-altmejd2019}{}}%
Altmejd A, Dreber A, Forsell E, Huber J, Imai T, Johannesson M, Kirchler M, Nave G, Camerer C (2019) Predicting the replicability of social science lab experiments. \emph{PLOS ONE} \textbf{14}:e0225826. doi:\href{https://doi.org/10.1371/journal.pone.0225826}{10.1371/journal.pone.0225826}

\leavevmode\vadjust pre{\hypertarget{ref-anderson2016}{}}%
Anderson CJ, Bahník Š, Barnett-Cowan M, Bosco FA, Chandler J, Chartier CR, Cheung F, Christopherson CD, Cordes A, Cremata EJ, Della Penna N, Estel V, Fedor A, Fitneva SA, Frank MC, Grange JA, Hartshorne JK, Hasselman F, Henninger F, Hulst M van der, Jonas KJ, Lai CK, Levitan CA, Miller JK, Moore KS, Meixner JM, Munafò MR, Neijenhuijs KI, Nilsonne G, Nosek BA, Plessow F, Prenoveau JM, Ricker AA, Schmidt K, Spies JR, Stieger S, Strohminger N, Sullivan GB, Aert RCM van, Assen MALM van, Vanpaemel W, Vianello M, Voracek M, Zuni K (2016) Response to {Comment} on {``{Estimating} the reproducibility of psychological science.''} \emph{Science} \textbf{351}:1037--1037. doi:\href{https://doi.org/10.1126/science.aad9163}{10.1126/science.aad9163}

@@

\leavevmode\vadjust pre{\hypertarget{ref-crump2013}{}}%
Crump MJC, McDonnell JV, Gureckis TM (2013) Evaluating amazon's mechanical turk as a tool for experimental behavioral research. \emph{{PLOS} {ONE}} \textbf{8}:e57410. doi:\href{https://doi.org/10.1371/journal.pone.0057410}{10.1371/journal.pone.0057410}

@@

\leavevmode\vadjust pre{\hypertarget{ref-hagger2016}{}}%
Hagger MS, Chatzisarantis NLD, Alberts H, Anggono CO, Batailler C, Birt AR, Brand R, Brandt MJ, Brewer G, Bruyneel S, Calvillo DP, Campbell WK, Cannon PR, Carlucci M, Carruth NP, Cheung T, Crowell A, De Ridder DTD, Dewitte S, Elson M, Evans JR, Fay BA, Fennis BM, Finley A, Francis Z, Heise E, Hoemann H, Inzlicht M, Koole SL, Koppel L, Kroese F, Lange F, Lau K, Lynch BP, Martijn C, Merckelbach H, Mills NV, Michirev A, Miyake A, Mosser AE, Muise M, Muller D, Muzi M, Nalis D, Nurwanti R, Otgaar H, Philipp MC, Primoceri P, Rentzsch K, Ringos L, Schlinkert C, Schmeichel BJ, Schoch SF, Schrama M, Schütz A, Stamos A, Tinghög G, Ullrich J, vanDellen M, Wimbarti S, Wolff W, Yusainy C, Zerhouni O, Zwienenberg M (2016) A multilab preregistered replication of the ego-depletion effect. \emph{Perspect Psychol Sci} \textbf{11}:546--573. doi:\href{https://doi.org/10.1177/1745691616652873}{10.1177/1745691616652873}

@@

\leavevmode\vadjust pre{\hypertarget{ref-openscienceconsortium2015}{}}%
Open Science Consortium (2015) \href{https://www.science.org/doi/full/10.1126/science.aac4716?casa_token=IJ35TwwlcjsAAAAA\%3AqiP68QbVAHleIg9zD3WugKWuV6Oa5rswS0VQnDsCq5I14ME4WIQabNGVD_T6SBSuAt6voVHNnWc0sw}{Estimating the reproducibility of psychological science}. \emph{Science}
